Trending assessment for nowcasts, measurements, and forecasts

### 2024-03-05 sections/03-trending.tex
Update on Overleaf.

As described in Section~\ref{sec:introduction}, we refer by trending to the consistency of an observed change with the change issued by another measurement, a nowcast, or a forecast.
In this section, we want to specify trending and define different methods of measuring and visualizing trending. 
In general, trending can be assessed for different lags $l$ separately, and the relevance of the considered lags has to be determined based on the question at hand.
Throughout the section, we call a measurement, nowcast, or forecast a \textit{prediction} or \textit{signal} whenever the application area is insignificant.
Before introducing measures to evaluate trending, we make remarks on the calculation of predicted change and notation in Section~\ref{subsec:notation}.
Section~\ref{subsec:trending-basics} introduces the basics of trending and visualizes trending in a synthetic example.
In Section~\ref{subsec:trending-measures}, various methods from other disciplines and tasks are applied and examined.
The approaches are extended to account for noise and non-systematic effects in Section~\ref{subsec:trending-noise}.
Section~\ref{subsec:trending-cond-prob} introduces a new graphical method for local trending assessment and briefly introduces bootstrap methods for computing confidence intervals for the measures in the previous sections.
Trending assessment for probabilistic nowcasts and forecasts is treated in Section~\ref{subsec:probabilistic}.

@@

As noted above, trending refers to observed and predicted changes for a time lag $l$.
The \textit{observed} change is straightforward to compute for all types of signals.
Let $\mathbf{y} = (y_t)_{t=0}^T$ denote the true values or gold standard measurements up to time $T$.
The sequence of observed change is then given by the differences of values in $\mathbf{y}$ with lag $l$, that is
\begin{equation}\label{eq:diffy}
    \diffylt = (y_{t+l} - y_t) \quad \text{for}\ t = 1, \dots, T - l.
\end{equation}

@@

For measurement, a test method is usually evaluated against a gold standard to test whether the test method could replace the gold standard.
If the test method is assessed to be consistent with the gold standard, the gold standard measurement values are not available to compute the predicted change.
Thus, the predicted change is computed on the basis of the test method's values at time $t$ and $t + l$.

Taking the three points together, the signal's predicted change is based on the type of signal.
To account for several signals for a target time $t$ at different issue times $\tau$, we denote those by $x_{t | \tau}$.
For nowcasting, $\tau$ is usually not considerably larger than $t$; while $t$ is usually at least $\tau$ in forecasting.
For measurements, this distinction is not necessary.
Table~\ref{tab:notation} summarizes the notation for the different signal types and the computation of the predicted change.

@@

Analyzing the number of points in the green quadrants versus the red quadrants is a standard approach in trending assessment of measurement data \parencite{Critchley2010,Saugel2015}.
It boils down to assessing the probability of trending $P(\diffxrv \diffyrv > 0)$, where $\diffyrv$ and $\diffxrv$ denote random variables for future incremental changes, respectively.
Note that $z_1 z_2 > 0$ if and only if $\sign(z_1) = \sign(z_2)$ ($z_1, z_2 \in \R \setminus \{ 0 \}$).
The standard estimator for $P(\diffxrv \diffyrv > 0)$ is
\begin{equation}
    \acc (\diffx, \diffy) \coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0}}{\card{T}}.\label{eq:acc}
\end{equation}
We refer to this estimator as the trending ratio of the prediction and set $\mathcal{T} = \{1, \dots, T-l\}$.
Visually, the measure computes the fraction of points in the upper right or lower left quadrant.
Such a $2 \times 2$ table of counts is called a contingency table.
It is often used in other scientific areas for evaluation, for example, in dichotomous forecasting or as a confusion matrix in classification analysis~\parencites(see, e.g., the introductions in)()[Ch. 4]{James2021}[Ch. 3]{Jolliffe2012}.
There, a wide range of other methods are usually used to analyze further characteristics of contingency tables.
Two simple measures that focus on specific areas of interest are the positive and negative trending ratios $\accp$ and $\accm$, respectively.
They are defined as
\begin{align}
    \accp (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0}} \label{eq:accp}\\
    \accm (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0}}\label{eq:accm}
\end{align}
In the classification context, these measures are known as positive or negative predictive value and hit rate or detection failure ratio in forecasting.
They give the probability of a correct prediction of the direction of change, given that the signal direction is positive or negative, respectively.
Therefore, they measure the trending ability of the prediction for positive and negative changes separately by cutting the considered data.
In terms of probability, the two measures give estimates of $P(\diffxrv \diffyrv > 0 | \diffxrv > 0)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0)$, respectively.

The use of these estimators could encourage predictors to exploit imbalances in the number of positive and negative changes. 
A large difference in $\sum_{t \in \mathcal{T}} \ind{\diffyt > 0}$ and $\sum_{t in \mathcal{T}} \ind{\diffyt > 0}$ is unlikely in the trending setting, since $\diffy$ is obtained from differencing time series data.
However, if the number of positive and negative $\diffy$ differs widely, the use of unbalanced-data-aware measures should be considered.
There are various adapted measures for unbalanced outcomes, for example, Cohen's $\kappa$ \parencite{Cohen1960} or those listed in \textcite[Table 3.3]{Jolliffe2012}.
Cohen's $\kappa$ is usually used to measure inter-rater agreement and takes into account the ratio of occurred agreement and the probability of agreement by chance.
Cohen's $\kappa$ reduces to rescaling the trending ratio in the case of a $2\times2$ table and balanced outcomes.

All measures can also be evaluated as a rolling estimate to detect changes in performance over time.
A rolling estimate is a sequence of estimates, each computed on a length-$w$-window of the data.
For the trending ratio, a rolling estimate with a backward-looking window is given by
\begin{equation*}
    \acc_t (\diffx, \diffy) \coloneqq \frac{\sum_{t^\star = t-w + 1}^{t} \ind{\diffxt[t^\star] \diffyt[t^\star] > 0}}{w} \quad t = w-1, \dots, T.\label{eq:acc_rolling}
\end{equation*}
Figure~\ref{fig:trending_ratio_time_series} depicts a rolling window estimate of the trending ratio for the simulated data of Figures~\ref{fig:trending_basic_4q_sample} and~\ref{fig:trending_basic_4q_sample_color}.
Thus, the yearly course of the prediction's trending ratio can be detected.
The trending ability of the signal has a strong yearly sinus-shaped seasonality with a trending ability peak after a quarter of a year and a low point after three quarters.
This seasonal behavior cannot be distinguished in the two four-quadrant plots or in the trending ratio.

\begin{figure}
    \centering
    \begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_sample_without_time}
\caption{Four-quadrant plot with simulated data.}\label{fig:trending_basic_4q_sample}
\end{subfigure}\hspace{0.01\textwidth}
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_sample_with_time}
\caption{The same data is colored according to the time index $t$, the greener, the later.}\label{fig:trending_basic_4q_sample_color}
\end{subfigure}\hspace{0.01\textwidth}
\begin{subfigure}[t]{.48\textwidth}
    \includegraphics{plots/illustrative_examples/trending_ratio_time_series.pdf}
    \caption{Rolling estimate of the trending ratio over time with window length 91. }\label{fig:trending_ratio_time_series}
    \end{subfigure}%
    \caption{Visualizations for data with a time-varying trending ratio. We defer information on the data generation process of the data in \ref{fig:trending_basic_4q_sample} and \ref{fig:trending_basic_4q_sample_color} to the appendix (see \ref{subsec:app-trending-data-generation}). The trending ratios on the whole data set are $\mu = 0.7577$. The strong yearly seasonality of the trending ratio becomes visible in Figure~\ref{fig:trending_ratio_time_series}. The green curve $k_t$ shows the probability that $\diffxt$ has the same sign as $\diffyt$ for each time step. The rolling estimates are delayed as the windows look backward.}
\end{figure}

In applications, signal data are often not available for all time steps, for example, due to technical problems or delays in data transfer (see the examples in Sections~\ref{sec:application-covid} and~\ref{sec:application-eda}).
We refer to time steps for which either signal or true values or both are unavailable as missing values.
Pairs of data with missing values can be excluded from the set $\mathcal{T}$ to calculate the measures. 
However, systematical missing values could lead to a biased estimate of the trending ratio, for example, if signal data is omitted in times of high change and thus uncertainty.
In this case, measures should be interpreted with caution, and
inspection of missing data should accompany applications with missing data, for example, through a visual assessment.
In the applications in Sections~\ref{sec:application-covid} and~\ref{sec:application-eda}, only a few missing values occur, and we provide information on the missing data by inspecting the missing data as lists.

@@

Using an exclusion area around the zero point is a straightforward and highly interpretable extension of the measures of Section~\ref{subsec:trending-measures}.
Points within that area are neither plotted in the four-quadrant plot nor included in the calculation of the measures.
The shape of the exclusion area can be chosen according to the noise characteristics of predictions and true values.
In general, it is likely that the prediction models have a noise component and should thus be part of the exclusion area.
We denote the measures of Equations~\eqref{eq:acc},~\eqref{eq:accp} and~\eqref{eq:accm} accounting for an exclusion area $e$ by
\begin{align}
    \acceps (\diffx, \diffy, e) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffx \diffy > 0} \ind{(\diffyt, \diffxt) \notin e}}{\sum_{t \in \mathcal{T}} \ind{(\diffyt, \diffxt) \notin e}}\label{eq:acceps}\\
    \accpeps (\diffx, \diffy, e) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0, , (\diffyt, \diffxt) \notin e}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0, , (\diffyt, \diffxt) \notin e}} \label{eq:accpeps}\\
    \accmeps (\diffx, \diffy, e) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0, (\diffyt, \diffxt) \notin e}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0, (\diffyt, \diffxt) \notin e}}\label{eq:accmeps}
\end{align}
The measures are then estimators for the probability of trending, given that the change is not in the exclusion area $e$, that is, for $\acceps$, $P(\diffxrv \diffyrv > 0 | \diffxrv \diffyrv \notin e)$.
The estimators accept various shapes of the exclusion area.
Figure~\ref{fig:trending_4q} visualizes different shapes of the exclusion area.
A rectangular exclusion area, $e = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \land (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ for $\varepsilon_x, \varepsilon_y > 0$, leaves out points that are small in both components and therefore likely to be driven by noise.
Points where at least the true value or signal is unlikely to be zero are not excluded.
In the example graph in Figure~\ref{fig:trending_basic_4q_excl_box}, only point 1 is excluded.
An exclusion area along one axis, for example, $e = \{(x, y) \in \R^2: (-\varepsilon_x \leq 0 \leq \varepsilon_x)\}$ for $\varepsilon_x > 0$, removes points in which one of the components could change sign by a small amount of noise.
This particularly suits signals where small amounts of noise are inevitable.
A cross-shaped exclusion area, $e = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \lor (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ for $\varepsilon_x, \varepsilon_y > 0$, along both axes accounts for the sign reversal in both components.
For these two methods, points 1 and 7 in Figure~\ref{fig:trending_basic_4q_excl_axis} or 1, 6, and 7 are excluded in Figure~\ref{fig:trending_basic_4q_excl_cross}, respectively.

@@

In most applications, the shape and size of the exclusion area can be chosen based on domain knowledge or expert opinions.
The size estimation can also be based on a proportion of the total variance or the total range of the data.
A third approach is to visualize the trending ratio for different values of $e$.
Section~\ref{sec:application} contains various examples of such a plot.
As a base approach, we suggest using no exclusion area in the four-quadrant plot, as the points do not complicate its interpretation, and a $\acceps$-over-$e$ plot to visualize the course of the trending ratio over different exclusion area sizes.

\subsection{Bootstrap confidence intervals and the conditional trending plot}\label{subsec:trending-cond-prob}

Confidence intervals can account for the estimation uncertainty of the measures above.
Bootstrap confidence intervals are a nonparametric technique based on resampling~\parencite[for introductions see][]{Hesterberg2011,Bittmann2021}.
New samples are drawn with replacement from the dataset and are not based on parametric assumptions as for classical confidence intervals.
The confidence interval is then computed based on the bootstrap samples.
We examine three methods for bootstrapping: the intuitive percentile and the more sophisticated basic and \ac{bca} method.
In the \textit{percentile} approach, the confidence interval for level $\alpha$ is built directly from the empirical distribution of the bootstrap estimators.
The \textit{basic} approach computes the confidence interval based on the non-bootstrap estimate using the bootstrapped quantile deviations~\parencite{Davison1997}.
The \ac{bca} method modifies the quantiles of the empirical bootstrap distribution by a bias and an acceleration parameter~\parencite{Efron1987}.
Typically, the percentile approach needs larger datasets and provides an easy and fast estimate, while the \ac{bca} is computationally expensive but requires smaller datasets for reasonable confidence intervals.
The basic approach balances these two aims.
We compare the approaches in a small synthetic data study on their small-dataset behavior and computation time in Appendix~\ref{sec:appendix}.
The \ac{bca} is the only method that holds the confidence level also for small datasets while increasing the computation time only moderately for larger datasets.
Therefore, we use the \ac{bca} method for confidence intervals in the applications in Section~\ref{sec:application}.

@@

A multivariate \acf{kde} facilitates a continuous estimation of $P(\diffxrv \diffyrv > 0 | \diffxrv = x)$ by estimating the components $f_{\diffxrv, \diffyrv}$ and $f_{\diffxrv}$ of
\begin{align*}
P(\diffxrv \diffyrv > 0 | \diffxrv = x) = \begin{cases}
                                              \int_{-\infty}^0 \frac{f_{\diffxrv, \diffyrv}(x, y)}{f_{\diffxrv}(x)} \ \textrm{d} \: y & \text{if } x < 0, \\
                                              \int_{0}^{\infty} \frac{f_{\diffxrv, \diffyrv}(x, y)}{f_{\diffxrv}(x)} \ \textrm{d} \: y & \text{if } x > 0, \\
\end{cases}
\end{align*}
for $x \neq 0$ through a \ac{kde}.
A comprehensive introduction to multivariate \ac{kde} can be found in \textcite{Gramacki2018}, and implementations are available in many programming languages~\parencite[e.g., for  Python in][]{Seabold2010}.
The \ac{kde} yields estimates for $P(\diffxrv \diffyrv > 0 | \diffxrv = x)$ for all values of $x \in \R$.
Multivariate \ac{kde} takes as modeling parameters a kernel and a bandwidth selector. 
The kernel is usually Gaussian, and there are multiple methods for selecting the bandwidth, each with its own advantages and disadvantages.
Note that for kernels with infinite support, for example, the Gaussian kernel, the plot should be limited to the core values of $\diffx$ without outliers to avoid estimates based on small subsets of the data.

>>>>

As described in Section~\ref{sec:introduction}, we refer by trending to the consistency of an observed change with the change issued by another measurement, a nowcast, or a forecast.
In this section, we want to specify trending and define different methods of measuring and visualizing trending.
\unsure{ist das Wort "observed" wirklich das was wir meinen? Stichwort: Rauschen}
In general, trending can be assessed for different lags $l$ separately, and the relevance of the considered lags has to be determined based on the question at hand.
Throughout the section, we call a measurement, nowcast, or forecast a \textit{prediction} or \textit{signal} whenever the application area is insignificant.
Before introducing measures to evaluate trending, we make remarks on the calculation of predicted change and notation in Section~\ref{subsec:notation}.
Section~\ref{subsec:trending-basics} introduces the basics of trending and visualizes trending in a synthetic example.
In Section~\ref{subsec:trending-measures}, various methods from other disciplines and tasks are applied and examined.
The approaches are extended to account for noise and non-systematic effects in Section~\ref{subsec:trending-noise}.
Section~\ref{subsec:trending-cond-prob} introduces a new graphical method for local trending assessment and briefly introduces bootstrap methods for computing confidence intervals for the measures in the previous sections.
Trending assessment for probabilistic nowcasts and forecasts is treated in Section~\ref{subsec:probabilistic}.

@@

As noted above, trending refers to observed and predicted changes for a time lag $l$.
The \textit{observed} change is straightforward to compute for all types of signals.
Let $\mathbf{y} = (y_t)_{t=0}^T$ denote the true values or gold standard measurements up to time $T$.
\unsure{Vielleicht schreiben wir hier noch ausfürhlich aus, was dieses y in drei Fällen sein sollte, oder ist das klar nach introduction?}
The sequence of observed change is then given by the differences of values in $\mathbf{y}$ with lag $l$, that is
\begin{equation}\label{eq:diffy}
    \diffylt = (y_{t+l} - y_t) \quad \text{for}\ t = 1, \dots, T - l.
\end{equation}

@@

For measurement, a test method is usually evaluated against a gold standard to test whether the test method could replace the gold standard.
If the test method is assessed to be consistent with the gold standard, the gold standard measurement values are not available to compute the predicted change. \unsure{irgendwie für mich nicht klar formuliert: man unteruscht ja bei meausurement nur ob die gemessenen Änderungen von den zwei Methoden konsistent sind oder?}
Thus, the predicted change is computed on the basis of the test method's values at time $t$ and $t + l$.

Taking the three points together, the signal's predicted change is based on the type of signal.
To account for several signals for a target time $t$ at different issue times $\tau$, we denote those by $x_{t | \tau}$.
For nowcasting, $\tau$ is usually not considerably larger than $t$; while $t$ is usually at least $\tau$ in forecasting. \unsure{Nowcast macht Schätzung i. d. R. so lange, bis der echte Wert bekannt wird oder?}
For measurements, this distinction is not necessary.
Table~\ref{tab:notation} summarizes the notation for the different signal types and the computation of the predicted change.

@@

Analyzing the number of points in the green quadrants versus the red quadrants is a standard approach in trending assessment of measurement data \parencite{Critchley2010,Saugel2015}. \unsure{Quelle-Fehler}
It boils down to assessing the probability of trending $P(\diffxrv \diffyrv > 0)$, where $\diffyrv$ and $\diffxrv$ denote random variables for future incremental changes, respectively.
Note that $z_1 z_2 > 0$ if and only if $\sign(z_1) = \sign(z_2)$ ($z_1, z_2 \in \R \setminus \{ 0 \}$).
The standard estimator for $P(\diffxrv \diffyrv > 0)$ is
\begin{equation}
    \acc (\diffx, \diffy) \coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0}}{\card{T}}.\label{eq:acc}
\end{equation}
We refer to this estimator as the trending ratio of the prediction and set $\mathcal{T} = \{1, \dots, T-l\}$.
Visually, the measure computes the fraction of points in the upper right or lower left quadrant.
Such a $2 \times 2$ table of counts is called a contingency table.
It is often used in other scientific areas for evaluation, for example, in dichotomous forecasting or as a confusion matrix in classification analysis~\parencites(see, e.g., the introductions in)()[Ch. 4]{James2021}[Ch. 3]{Jolliffe2012}.
There, a wide range of other methods are usually used to analyze further characteristics of contingency tables.
Two simple measures that focus on specific areas of interest are the positive and negative trending ratios $\accp$ and $\accm$, respectively.
They are defined as
\begin{align}
    \accp (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0}} \label{eq:accp}\\
    \accm (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0}}\label{eq:accm}
\end{align}
In the classification context, these measures are known as positive or negative predictive value and hit rate or detection failure ratio in forecasting.
They give the probability of a correct prediction of the direction of change, given that the signal direction is positive or negative, respectively.
Therefore, they measure the trending ability of the prediction for positive and negative changes separately by cutting the considered data.
In terms of probability, the two measures give estimates of $P(\diffxrv \diffyrv > 0 | \diffxrv > 0)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0)$, respectively.

The use of these estimators could encourage predictors to exploit imbalances in the number of positive and negative changes. 
A large difference in $\sum_{t \in \mathcal{T}} \ind{\diffyt > 0}$ and $\sum_{t \in \mathcal{T}} \ind{\diffyt > 0}$ is unlikely in the trending setting, since $\diffy$ is obtained from differencing time series data. \unsure{hier stimmt was nicht}
However, if the number of positive and negative $\diffy$ differs widely, the use of unbalanced-data-aware measures should be considered.
There are various adapted measures for unbalanced outcomes, for example, Cohen's $\kappa$ \parencite{Cohen1960} or those listed in \textcite[Table 3.3]{Jolliffe2012}.
Cohen's $\kappa$ is usually used to measure inter-rater agreement and takes into account the ratio of occurred agreement and the probability of agreement by chance.
Cohen's $\kappa$ reduces to rescaling the trending ratio in the case of a $2\times2$ table and balanced outcomes.

All measures can also be evaluated as a rolling estimate to detect changes in performance over time.
A rolling estimate is a sequence of estimates, each computed on a length-$w$-window of the data.
For the trending ratio, a rolling estimate with a backward-looking window is given by
\begin{equation*}
    \acc_t (\diffx, \diffy) \coloneqq \frac{\sum_{t^\star = t-w + 1}^{t} \ind{\diffxt[t^\star] \diffyt[t^\star] > 0}}{w}, \quad t = w-1, \dots, T.\label{eq:acc_rolling}
\end{equation*}
Figure~\ref{fig:trending_ratio_time_series} depicts a rolling window estimate of the trending ratio for the simulated data of Figures~\ref{fig:trending_basic_4q_sample} and~\ref{fig:trending_basic_4q_sample_color}.
Thus, the yearly course of the prediction's trending ratio can be detected.
The trending ability of the signal has a strong yearly sinus-shaped seasonality with a trending ability peak after a quarter of a year and a low point after three quarters.
This seasonal behavior cannot be distinguished in the two four-quadrant plots or in the trending ratio.

\begin{figure}
    \centering
    \begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_sample_without_time}
\caption{Four-quadrant plot with simulated data.}\label{fig:trending_basic_4q_sample}
\end{subfigure}\hspace{0.01\textwidth}
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_sample_with_time}
\caption{The same data is colored according to the time index $t$, the greener, the later.}\label{fig:trending_basic_4q_sample_color}
\end{subfigure}\hspace{0.01\textwidth}
\begin{subfigure}[t]{.48\textwidth}
    \includegraphics{plots/illustrative_examples/trending_ratio_time_series.pdf}
    \caption{Rolling estimate of the trending ratio over time with window length 91. }\label{fig:trending_ratio_time_series}
    \end{subfigure}%
    \caption{Visualizations for data with a time-varying trending ratio. We defer information on the data generation process of the data in \ref{fig:trending_basic_4q_sample} and \ref{fig:trending_basic_4q_sample_color} to the appendix (see \ref{subsec:app-trending-data-generation}). The trending ratios on the whole data set are $\mu = 0.7577$. The strong yearly seasonality of the trending ratio becomes visible in Figure~\ref{fig:trending_ratio_time_series}. The green curve $k_t$ shows the probability that $\diffxt$ has the same sign as $\diffyt$ for each time step. The rolling estimates are delayed as the windows look backward.}
\end{figure} 
\unsure{was ist $\mu_1$ bei dem Plot, hinschreiben}
In applications, signal data are often not available for all time steps, for example, due to technical problems or delays in data transfer (see the examples in Sections~\ref{sec:application-covid} and~\ref{sec:application-eda}).
We refer to time steps for which either signal or true values or both are unavailable as missing values.
Pairs of data with missing values can be excluded from the set $\mathcal{T}$ to calculate the measures. 
However, systematical missing values could lead to a biased estimate of the trending ratio, for example, if signal data is omitted in times of high change and thus uncertainty.
In this case, measures should be interpreted with caution, and
inspection of missing data should accompany applications with missing data, for example, through a visual assessment.
In the applications in Sections~\ref{sec:application-covid} and~\ref{sec:application-eda}, only a few missing values occur, and we provide information on the missing data by inspecting the missing data as lists.

@@

Using an exclusion area around the zero point is a straightforward and highly interpretable extension of the measures of Section~\ref{subsec:trending-measures}.
Points within that area are neither plotted in the four-quadrant plot nor included in the calculation of the measures.
The shape of the exclusion area can be chosen according to the noise characteristics of predictions and true values.
In general, it is likely that the prediction models have a noise component and should thus be part of the exclusion area.\unsure{vielleicht eine große Buchstabe für exclusion area?}
We denote the measures of Equations~\eqref{eq:acc},~\eqref{eq:accp} and~\eqref{eq:accm} accounting for an exclusion area $e$ by
\begin{align}
    \acceps (\diffx, \diffy, e) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffx \diffy > 0} \ind{(\diffyt, \diffxt) \notin e}}{\sum_{t \in \mathcal{T}} \ind{(\diffyt, \diffxt) \notin e}}\label{eq:acceps}\\
    \accpeps (\diffx, \diffy, e) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0, , (\diffyt, \diffxt) \notin e}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0, (\diffyt, \diffxt) \notin e}} \label{eq:accpeps}\\
    \accmeps (\diffx, \diffy, e) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0, (\diffyt, \diffxt) \notin e}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0, (\diffyt, \diffxt) \notin e}}\label{eq:accmeps}
\end{align}
The measures are then estimators for the probability of trending, given that the change is not in the exclusion area $e$, that is, for $\acceps$, $P(\diffxrv \diffyrv > 0 | \diffxrv \diffyrv \notin e)$.
The estimators accept various shapes of the exclusion area.
Figure~\ref{fig:trending_4q} visualizes different shapes of the exclusion area.
A rectangular exclusion area, $e = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \land (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ ,for $\varepsilon_x, \varepsilon_y > 0$, leaves out points that are small in both components and therefore likely to be driven by noise.
Points where at least the true value or signal is unlikely to be zero are not excluded.
In the example graph in Figure~\ref{fig:trending_basic_4q_excl_box}, only point 1 is excluded.
An exclusion area along one axis, for example, $e = \{(x, y) \in \R^2: (-\varepsilon_x \leq 0 \leq \varepsilon_x)\}$ for $\varepsilon_x > 0$, removes points in which one of the components could change sign by a small amount of noise.
This particularly suits signals where small amounts of noise are inevitable.
A cross-shaped exclusion area, $e = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \lor (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ for $\varepsilon_x, \varepsilon_y > 0$, along both axes accounts for the sign reversal in both components.
For these two methods, points 1 and 7 in Figure~\ref{fig:trending_basic_4q_excl_axis} or 1, 6, and 7 are excluded in Figure~\ref{fig:trending_basic_4q_excl_cross}, respectively.

@@

In most applications, the shape and size of the exclusion area can be chosen based on domain knowledge or expert opinions.
The size estimation can also be based on a proportion of the total variance or the total range of the data.
A third approach is to visualize the trending ratio for different values of $e$.
Section~\ref{sec:application} contains various examples of such a plot.
As a base approach, we suggest using no exclusion area in the four-quadrant plot, as the points do not complicate its interpretation, and a $\acceps$-over-$e$ plot to visualize the course of the trending ratio over different exclusion area sizes.\unsure{"as the points do not complicate its interpretation" ist das hier die "richtige" Begründung, dass wir kein exlclusion area betrachten?}

\subsection{Bootstrap confidence intervals and the conditional trending plot}\label{subsec:trending-cond-prob}
\unsure{Vielleicht doch zwei subsections?}
Confidence intervals can account for the estimation uncertainty of the measures above.
Bootstrap confidence intervals are a nonparametric technique based on resampling~\parencite[for introductions see][]{Hesterberg2011,Bittmann2021}.
New samples are drawn with replacement from the dataset and are not based on parametric assumptions as for classical confidence intervals.
The confidence interval is then computed based on the bootstrap samples.
We examine three methods for bootstrapping: the intuitive percentile and the more sophisticated basic and \ac{bca} method.
In the \textit{percentile} approach, the confidence interval for level $\alpha$ is built directly from the empirical distribution of the bootstrap estimators.
The \textit{basic} approach computes the confidence interval based on the non-bootstrap estimate using the bootstrapped quantile deviations~\parencite{Davison1997}.
The \ac{bca} method modifies the quantiles of the empirical bootstrap distribution by a bias and an acceleration parameter~\parencite{Efron1987}.
Typically, the percentile approach needs larger datasets and provides an easy and fast estimate, while the \ac{bca} is computationally expensive but requires smaller datasets for reasonable confidence intervals.
The basic approach balances these two aims.
We compare the approaches in a small synthetic data study on their small-dataset behavior and computation time in Appendix~\ref{sec:appendix}.
The \ac{bca} is the only method that holds the confidence level also for small datasets while increasing the computation time only moderately for larger datasets.
Therefore, we use the \ac{bca} method for confidence intervals in the applications in Section~\ref{sec:application}.

@@

A multivariate \acf{kde} facilitates a continuous estimation of $P(\diffxrv \diffyrv > 0 | \diffxrv = x)$ by estimating the components $f_{\diffxrv, \diffyrv}$ and $f_{\diffxrv}$ of
\begin{align*}
P(\diffxrv \diffyrv > 0 | \diffxrv = x) = \begin{cases}
                                              \int_{-\infty}^0 \frac{f_{\diffxrv, \diffyrv}(x, y)}{f_{\diffxrv}(x)} \ \textrm{d} \: y & \text{if } x < 0, \\
                                              \int_{0}^{\infty} \frac{f_{\diffxrv, \diffyrv}(x, y)}{f_{\diffxrv}(x)} \ \textrm{d} \: y & \text{if } x > 0, \\
\end{cases}
\end{align*}
for $x \neq 0$ through a \ac{kde}.

A comprehensive introduction to multivariate \ac{kde} can be found in \textcite{Gramacki2018}, and implementations are available in many programming languages~\parencite[e.g., for  Python in][]{Seabold2010}.
The \ac{kde} yields estimates for $P(\diffxrv \diffyrv > 0 | \diffxrv = x)$ for all values of $x \in \R$.
Multivariate \ac{kde} takes as modeling parameters a kernel and a bandwidth selector. 
The kernel is usually Gaussian, and there are multiple methods for selecting the bandwidth, each with its own advantages and disadvantages.
Note that for kernels with infinite support, for example, the Gaussian kernel, the plot should be limited to the core values of $\diffx$ without outliers to avoid estimates based on small subsets of the data.
